---
title: "Regression Models Project"
output: pdf_document
geometry: margin=0.5in
---

 ```{r global_options, include=FALSE}
 knitr::opts_chunk$set(fig.width=8, fig.height=4, fig.align="center", fig.path='figs/',
                       echo=TRUE, warning=FALSE, message=FALSE)
 ```


### Executive Summary

In this project, we're asked to use the *mtcars* dataset and write up an analysis to
answer two main questions:

* "Is an automatic or manual transmission better for MPG"
* "Quantify the MPG difference between automatic and manual transmissions"

To answer these questions we'll build a number of linear regression models in increased complexity
and compare them in terms of their performance. As a result of this study we find that manual
transmission cars in this dataset have better *mpg* by about **2** miles/gallon compared to automatic 
transmission cars. However this difference in performance is found not to be statistically significant
based on the final model used here.

### Analysis

Let us begin by investigating the dataset first:

```{r}
dim(mtcars)
head(mtcars,2)
```

As can be seen, there are a total of 32 observables (cars) that have 11 measurements.
One can simply do `?mtcars` in the console to see what each measurement corresponds to, 
and what their units are. For the sake of bervity, we'll not do this here.

In order to tackle the afore mentioned questions, let us build a few linear models, in increasing
complexity, following the course contet:

```{r cache=TRUE}
model1 <- lm(mpg ~ factor(am) - 1, data = mtcars)
model2 <- update(model1, . ~ . + wt)
model3 <- update(model2, . ~ . + factor(cyl))
model4 <- update(model3, . ~ . + hp)
model5 <- update(model4, . ~ . + disp)
```

Now, let's look at the coefficients of the first model:

```{r echo=FALSE}
coef(model1)
```

Here the outcome (*mpg*) is explained by a single regressor (*am* - transmission type) 
that is a factor variable. **-1** in the formula means we don't want to have the constant term.
As a result, we end up with two coefficients, one corresponding to the automatic transmission (0),
and the other to the manual transmission (1). The numerical values of the coefficients 
show the median *mpg* for each transmission type, namely **17** miles/gallon for the automatic 
transmission and **24** miles/gallon for the manual transmission. 
This can be verified looking at the first box plot in the Appendix.

Now, the question is "Is this the end of the story? Do manual transmission cars significantly outperform
automatic counterparts based on this dataset?". If one looks at the correlation matrix (second figure in the 
Appendix), which shows how much the variables are correlated to each other in the dataset, 
s/he immediately sees there are other variables that are more correlated to the *mpg* than the transmission type. 
This means that if we add them to our model we have a better chance to describe the variability in *mpg*.
Let's use ANOVA to see how the other models, where we add other variables, compare to the first one:

```{r echo=FALSE}
anova(model1, model2, model3,model4,model5)
```

As one can see, *Model 4*, which includes the weight, cylinder type, and horsepower in addition to
the transmission type performs the best. Adding displacement doesn't gain us much since it
is highly correlated to the other regressors. Let's look at the summary of this model:

```{r echo=FALSE}
summary(model4)
```

In this model, one can see that the cylinder type, horsepower, and the weight are all
negatively correlated to the milage (i.e. the coefficients are negative). The coefficient
of the weight is **-2.5**, which means that if all other varibles are kept fixed, the *mpg* 
will go down by **2.5** gallons for each additional **1000** lbs in weight. The R-squared value of 
this model tells us that it explains **99%** of the variation in *mpg*, which is very good.
A few diagnostic plots for this model, which don't show any striking issues, 
are shown in the third figure in the Appendix.
Now let's see the confidence intervals for the predicted milage for a hypothetical car 
that has the mean weight and horsepower as of the original dataset, and 6 cylinders:

```{r}
# Predict
model4.prediction <- predict(model4, 
                             data.frame(am = c(0,1), wt = mean(mtcars$wt), cyl = c(6), hp = mean(mtcars$hp)), 
                             interval="confidence")
print(model4.prediction)
```

### Conclusions

What we see is that if our hypothetical car has automatic transmission it has an
expected *mpg* of about **18** miles/gallon, and if it has manual transmission about **20** miles/gallon.
However, the **95%** confidence level intervals are not exclusive, i.e. although the manual 
transmission looks to have better milage (about **2** miles per gallon - which can be verified
by the fitted model coefficients for the transmission type), it's not statistically significant.

### Appendix

```{r echo=FALSE}
# Look at a few distributions in ggplot2
require(ggplot2)

# Add a simple factor variable for the transmission
mtcars.custom <- mtcars
mtcars.custom$transmission <- factor( mtcars$am, labels = c("Automatic","Manual") )

# MPG by Transmission
plot <- ggplot(mtcars.custom, aes(x = transmission, y = mpg, fill = transmission)) + geom_boxplot() 
plot <- plot + scale_x_discrete(name = "Transmission Type") 
plot <- plot + scale_y_continuous(name = "Milase per Gallon")
plot <- plot + ggtitle("Boxplot of Miles per Gallon by Transmission Type")
plot <- plot + theme(plot.title = element_text(hjust = 0.5))
plot <- plot + guides(fill=FALSE)
print(plot)

# Get the correlation plot
library(reshape2)
cormat <- round(cor(mtcars),2)
plot <- ggplot(melt(cormat), aes(x=Var1, y=Var2, fill=value)) + geom_tile()
plot <- plot + geom_text(aes(Var2, Var1, label = value), color = "black", size = 4)
plot <- plot + ggtitle("Correlation Matrix")
plot <- plot + scale_fill_gradient2(midpoint=0, low="blue", mid="white", high="red",
                                    limit = c(-1,1), space ="Lab", name = "Corr. Coeff.")
plot <- plot + theme( plot.title = element_text(hjust = 0.5),
                      axis.title.x = element_blank(), 
                      axis.title.y = element_blank())
print(plot)

# Model 4 Diagnostic Plots
library(ggfortify)
autoplot(model4, label.size = 3)

# Weight by Transmission
plot <- ggplot(mtcars.custom, aes(x = transmission, y = wt, fill = transmission)) + geom_boxplot() 
plot <- plot + scale_x_discrete(name = "Transmission Type") 
plot <- plot + scale_y_continuous(name = "Weight [1000 lbs]")
plot <- plot + ggtitle("Boxplot of Weight by Transmission Type")
plot <- plot + theme(plot.title = element_text(hjust = 0.5))
plot <- plot + guides(fill=FALSE)
print(plot)

# MPG by Weight
plot <- ggplot(mtcars.custom, aes(x = wt, y = mpg, color = transmission)) + geom_point(size = 3)
plot <- plot + scale_x_continuous(name = "Weight [1000 lbs]") 
plot <- plot + scale_y_continuous(name = "Miles per Gallon")
plot <- plot + ggtitle("Scatter plot of Miles per Gallon by Weight")
plot <- plot + theme(plot.title = element_text(hjust = 0.5))
plot <- plot + labs(colour = "Transmission Type")
plot <- plot + geom_smooth(method="lm")
print(plot)

# # MPG by Weight - Same Slope
# mtcars.custom$pred <- predict(model2)
# plot <- ggplot(mtcars.custom, aes(x = wt, y = mpg, color = transmission)) + geom_point(size = 3)
# plot <- plot + scale_x_continuous(name = "Weight [1000 lbs]")
# plot <- plot + scale_y_continuous(name = "Miles per Gallon")
# plot <- plot + ggtitle("Scatter plot of Miles per Gallon by Weight")
# plot <- plot + theme(plot.title = element_text(hjust = 0.5))
# plot <- plot + labs(colour = "Transmission Type")
# plot <- plot + geom_line(mapping=aes(y=pred))
# print(plot)

# MPG by Cylinder
plot <- ggplot(mtcars.custom, aes(x = cyl, y = mpg, color = transmission)) + geom_point(size = 3)
plot <- plot + scale_x_continuous(name = "Cylinders") 
plot <- plot + scale_y_continuous(name = "Miles per Gallon")
plot <- plot + ggtitle("Scatter plot of Miles per Gallon by Cylinders")
plot <- plot + theme(plot.title = element_text(hjust = 0.5))
plot <- plot + labs(colour = "Transmission Type")
plot <- plot + geom_smooth(method="lm")
print(plot)
```